---
title: "COVID-19 Cases and Average Annual Out of Pocket Healthcare Costs"
author: "Ben and Emma"
date: "2024-12-10"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
pacman::p_load(tidyverse, janitor, urbnmapr, stringr, robustbase, car, jtools)


```

### Introduction

The COVID-19 pandemic has been a research interest for finance and health researchers alike since the first wave of cases over four years ago. Early works have predicted how the pandemic would affect the healthcare industry, particularly with revenue loss associated with travel/social distancing restrictions (Barnett et al., 2020). As health professionals have gained a better understanding of the pandemic, researchers from the Centers for Disease Control and Prevention (CDC) have pursued studies on the persistent health impacts associated with positive COVID-19 tests (Wanga et al., 2021). Other investigations of the pandemic's financial impact in the years to follow focus primarily on how the costs related to COVID-19 might affect the profitability of hospitals (Li et al., 2023), yet this top-down macroeconomic approach to evaluating the financial burden of the pandemic potentially neglects the typical household impact that a high incidence of COVID-19 cases might pose. Studies into the typical cost of a COVID-19-related hospital visit give insight into the direct impact of the pandemic on insured individuals (Pike et al., 2023), but focusing on only insured individuals misrepresents many localities where COVID-19 cases were high and insurance-covered populations were low (McCain et al., 2022). Furthermore, the direct cost of hospital visits from COVID-19 is less telling of the aforementioned longitudinal effects that populations with high COVID-19 rates might experience. That is why this research project will focus on examining COVID-19 cases from 2021 and 2022 and their impact on out-of-pocket healthcare costs in 2022.

We decided to only focus on counties in Georgia in order to keep the scope of the project manageable and not muddy the analysis with too broad of a research question. Georgia was a state at the forefront of COVID-19 news, especially at the beginning of the pandemic. Governor Brian Kemp was one of the last governors to enact a statewide shelter in place order at the very beginning of the pandemic and one of the first to reopen businesses towards the end of April, 2020 (Cillizza, 2020). Additionally, Georgia ranks among the worst in regards to healthcare access. As of 2023, 15.6% of adults in Georgia went without care due to how much it would cost compared to the national average of 11.7% (The Commonwealth Fund, 2023). Georgia also had the third highest percentage of residents who lack health insurance coverage compared to other states (Masterson, 2023), which means more Georgia residents would have to pay out of pocket for healthcare costs compared to residents in a majority of other states. Keeping this in mind, it is important to understand how the COVID-19 pandemic has affected out of pocket healthcare costs in Georgia specifically due to the state of healthcare access in the state already.

To complete this analysis, we utilized parts of three different datasets as we were not able to find one dataset that included all of the necessary information to answer our research question: **What is the relationship between COVID-19 cases and Out of Pocket Healthcare Costs in Georgia, USA?**

#### **Outcome Variable: Out of Pocket Healthcare Costs**

One of the main variables we are interested in is out-of-pocket healthcare costs. We downloaded the data from [Policy Map](https://www.policymap.com/), but this information is from the Department of Health and Human Services' Agency for Healthcare Research and Quality (AHRQ) Medical Expenditure Panel Survey (2022) with demographic and health data from the U.S. Census American Community Survey (2018-2022). [Policy Map](https://www.policymap.com/) contains data pertaining to COVID-19 data on county, state/territory, and national levels in the United States. The website allows users to choose their denomination of interest and in some cases, their time frame of interest. We chose this site to be able to see county-wide data in Georgia and to have better control over the time frame of interest for our data, in this case 2022.

Below, we are downloading the dataset and selecting the information we need in order to do our analysis. This includes filtering the data to only include information from Georgia, renaming the variables to more intuitive names, and removing variables we are not interested in. This variable will ultimately describe the average annual out of pocket healthcare cost per person for each county in Georgia in 2022. We chose 2022 specifically in order to view the longitudinal effects of the pandemic (i.e. one year after peak COVID-19 case incidence rate) rather than the immediate effect of the pandemic, which might include observations of healthcare costs due primarily to COVID-19 itself (i.e. 2020).

```{r cleaning costs}

avg_healthcare_costs_2022 <- read_csv("./data/healthcarecosts2022.csv") %>% 
  row_to_names(row_number = 1) %>% 
  clean_names() %>%
  filter(sitsin_state == "GA") %>%
  rename(county = geo_id_name,
         avg_cost_2022 = totslf_avg) %>%
  mutate(fips = as.numeric(geo_id),
         avg_cost_2022=as.numeric(avg_cost_2022)) %>%
  select("county", "fips", "avg_cost_2022")
head(avg_healthcare_costs_2022)

```

#### **Independent Variable: COVID-19 Cases**

The other main data we are interested in exploring is related to COVID-19 cases. This data comes from journalists at the New York Times who pieced together this information from a variety of county, state, and national government sources. This data does have some limitations; due to the rapidly changing nature of the COVID-19 pandemic, officials did not always report data in the most consistent way. However, this would be true of almost any source reporting on COVID-19 cases.

The variable we are particularly interested in gathering from this data is the number of new COVID-19 cases per day, which includes both confirmed and probable cases. There are also limitations with this data as the New York Times prioritized the accuracy of cumulative case counts over new daily cases, which causes some of the new daily cases to appear anomalous as it is derived from differences in cumulative cases from one day to another. Largely, the observations in this dataset provide an accurate estimate for the trends in COVID-19 case incidence in 2021 and 2022.

(If you are interested in learning more about this dataset, the New York Times has published a [GitHub](https://github.com/nytimes/covid-19-data?tab=readme-ov-file#readme) with lots of data collected during the COVID-19 pandemic. The data used in our analysis comes specifically from the [Rolling Averages](https://github.com/nytimes/covid-19-data/tree/master/rolling-averages) file).

The reason we chose this variable to focus on rather than a variable describing the cumulative cases per day is due to the nature of our out-of-pocket healthcare cost variable. The out-of-pocket healthcare cost variable is an annual average. We wanted to make the COVID-19 case variable comparable, which means we would also need to have one observation per county per year. The alternative was choosing one specific time period to compare our COVID-19 case data, which was a route we considered, but ultimately decided against since any small discrete time period (such as a week) introduces potential confounding on a county-to-county basis; we have no way to confirm that we can make a direct comparison between counties for any one week/month, but on a larger scale (year), we can be more confident of an equivalent comparison. In this way, we are essentially increasing the sample size of days in trying to get a more accurate number for average daily COVID case rate over the span of the year. If we utilized a cumulative variable, when we would average over the year, we would be counting the same case many times (since it includes the observations from previous days), but since we are focusing on new cases per day, we can get the average daily new cases per county. This value would make it easier to compare between counties with widely different populations, which we will discuss later.

Below, we are downloading the data and selecting the information needed for our analysis. This includes filtering so we only have the data for Georgia as well as standardizing the names of variables. In order for us to later make an averaged variable, it is necessary to have one row per county rather than one row per date.

```{r cleaning 2021 cases}

covidcases_2021 <- read_csv("./data/us-counties-2021.csv") %>% 
  filter(state == "Georgia" & county != "Unknown") %>% 
  select("date", "geoid", "county", "cases") %>%
  pivot_wider(names_from = "date",
              values_from = "cases") %>%
  mutate(fips = as.numeric(str_remove(geoid, "USA-"))) %>% 
  select(-"geoid")
head(covidcases_2021)

```


```{r cleaning 2022 cases}

covidcases_2022 <- read_csv("./data/us-counties-2022.csv") %>% 
filter(state == "Georgia" & county != "Unknown") %>% 
  select("date", "geoid", "county", "cases") %>%
  pivot_wider(names_from = "date",
              values_from = "cases") %>%
  mutate(fips = as.numeric(str_remove(geoid, "USA-"))) %>% 
  select(-"geoid")
head(covidcases_2022)

```

##### **Population**

Lastly, we decided that in order for there to be a fair comparison, we needed to account for population in regard to COVID-19 cases. This last dataset we are utilizing is from the [United States Census Bureau](https://www.census.gov/data/datasets/time-series/demo/popest/2020s-counties-total.html). This contains population estimates for every county in Georgia based on the 2020 census and accounting for births, deaths, and immigration information per year. This is necessary to be able to compare counties like Gwinnett and Chattahoochee, where Gwinnett has nearly 10 times the population of Chattahoochee and would obviously have more COVID-19 cases as a result. This incongruence in population is what motivated our standardization of average daily COVID case rate to be considered per 100K people   

Below, we are downloading the data and selecting the variables of interest. This includes filtering for only data from Georgia counties and again standardizing the names of the variables.

```{r cleaning population}

county_population <- read_csv("./data/population.csv") %>% 
  clean_names() %>%
  filter(stname == "Georgia" & ctyname != "Georgia") %>%
  mutate(county = sprintf("%03d", as.numeric(county))) %>%
  unite(col = "fips", state, county, sep = "") %>%
  mutate(county = str_remove(ctyname, " County"),
         fips = as.numeric(fips)) %>%
  select("fips", "county", "popestimate2021", "popestimate2022")
head(county_population)

```

---

### Creating the Dataset

The three datasets we downloaded above and cleaned are now ready to be combined to be utilized in our exploratory analysis. Below, we are combining the COVID-19 cases for both 2021 and 2022 with the county population data. Here we are creating two new variables for each year in order to create variables that contain the average daily new COVID-19 cases for each year and standardize that data to the average per 100,000 people.

```{r joining cases and population, message=FALSE}

covidcases_population <- covidcases_2021 %>% 
  left_join(covidcases_2022, by = c("county", "fips")) %>% 
  left_join(county_population, by = c("county", "fips")) %>% 
  rowwise() %>% 
  mutate(avg_daily_cases_2021 = sum(c_across(starts_with("2021")), na.rm = T) / 365,
         avg_daily_cases_2022 = sum(c_across(starts_with("2022")), na.rm = T) / 365) %>% 
  mutate(cases_per_100k_2021 = (avg_daily_cases_2021/popestimate2021)*100000,
         cases_per_100k_2022 = (avg_daily_cases_2022/popestimate2022)*100000) %>% 
  select("county", "fips", "avg_daily_cases_2021", "avg_daily_cases_2022", "popestimate2021", "popestimate2022", "cases_per_100k_2021", "cases_per_100k_2022")
head(covidcases_population)

```

Lastly, we are joining the above dataset with our out-of-pocket healthcare data to ultimately have a dataset containing both of our primary interests: COVID-19 cases in 2021 and 2022 and out-of-pocket healthcare costs per county in Georgia. We are also adding a rank variable that we will explain at length when we map this data onto Georgia's county map.

```{r joining cases and cost}

covid_healthcarecosts <- avg_healthcare_costs_2022 %>% 
  left_join(covidcases_population, by = c("county", "fips")) %>%
  mutate(rank2021case=rank(-cases_per_100k_2021, ties.method="first"),
         rank2022case=rank(-cases_per_100k_2022, ties.method="first"),
         rank2022cost=rank(-avg_cost_2022, ties.method="first"))
head(covid_healthcarecosts)

```

### Codebook

Below is an explanation of the variables in our dataset for the exploratory analysis.

```{r codebook, echo = FALSE, results = TRUE}

Variable  <- c("county","fips","average annual healthcare cost", "2021 average daily cases", "2022 average daily cases", "2021 population estimate", "2022 population estimate", "2021 cases per 100K", "2022 cases per 100K", "2021 case rank", "2022 case rank", "2022 cost rank")
Definition <- c("county data was collected in", "unique code given to each county in the US", "annual out of pocket healthcare cost per person in 2022 measured in US dollars", "average daily new COVID-19 cases per county in 2021", "average daily new COVID-19 cases per county in 2022", "estimated population based on the 2020 census in 2021", "estimated population based on the 2020 census in 2022", "average daily new COVID-19 cases per 100,000 people in 2021", "average daily new COVID-19 cases per 100,000 people in 2022", "rank of each county with one being the highest cases per 100k in 2021", "rank of each county with one being the highest cases per 100k in 2022", "rank of each county with one being the highest out of pocket healthcare cost in 2022")
d <- data.frame(Variable,Definition)
knitr::kable(d)

```


---

### Preliminary Data Analysis

To start our data analysis, we first wanted to observe the spread and central tendency measures of our key variables.

#### **Outcome: Out of Pocket Healthcare Costs in 2022**

```{r cost summary, include = FALSE}

cost_sum <- summary(covid_healthcarecosts$avg_cost_2022)

```


```{r cost hist, echo = FALSE, results = TRUE}

ggplot(data = covid_healthcarecosts, mapping = aes(x = avg_cost_2022)) +
  geom_histogram(fill = "white", color = "black", bins = 8) +
  labs(title = "Distribution of Average Annual Out of Pocket \nHealthcare Costs per Person",
       subtitle = "Georgia, USA",
       x = "Average Annual Out of Pocket Healthcare Costs per Person in US$",
       y = "Number of Counties") +
  theme_minimal() +
  theme(axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold", size = 12),  # Larger y-axis labels
    axis.title = element_text(face = "bold", size = 14),  # Larger axis titles
    panel.grid.major = element_line(color = "lightgray"),  # Add light gray gridlines
    panel.background = element_rect(fill = "aliceblue"),  # Add light blue background
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5))


```

In 2022, across Georgia, the mean out of pocket healthcare cost per individual was approximately $`r round(cost_sum[4],0)` with a minimum of $`r cost_sum[1]` and a maximum of $`r cost_sum[6]` spend per person. The median was $`r cost_sum[3]`. With the median slightly less than the mean, this data is slightly right skewed as also shown on the histogram above.

#### **Independent Variables**

For our independent variables we also wanted to analyze their correlation with the outcome variable of average annual out of pocket healthcare costs per person.

##### **New Daily COVID-19 Cases per 100K in 2021**

```{r cases 2021 summary, include = FALSE}

cases2021_sum <- summary(covid_healthcarecosts$cases_per_100k_2021)

```


```{r cases 2021 histogram, echo = FALSE, results = TRUE}

ggplot(data = covid_healthcarecosts, mapping = aes(x = cases_per_100k_2021)) +
  geom_histogram(fill = "white", color = "black", bins = 20) +
  labs(title = "Distribution of Average New COVID-19 Cases \nper Day per 100,000 People",
       subtitle = "Georgia, USA - 2021",
       x = "",
       y = "Number of Counties") +
  theme_minimal() +
  theme(axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold", size = 12),  # Larger y-axis labels
    axis.title = element_text(face = "bold", size = 14),  # Larger axis titles
    panel.grid.major = element_line(color = "lightgray"),  # Add light gray gridlines
    panel.background = element_rect(fill = "aliceblue"),  # Add light blue background
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

```

In 2021, across Georgia, the mean average new COVID-19 cases per day per 100,000 people for all of Georgia in 2021 was `r round(cases2021_sum[4], 1)` with a minimum of `r round(cases2021_sum[1], 1)` new cases per day per 100,000 people and a maximum of approximately `r round(cases2021_sum[6], 1)` new cases per day per 100,000 people. The median is approximately `r round(cases2021_sum[3], 1)` new cases per day per 100,000 people. Because the median is slightly smaller than the mean, this data is very slightly right skewed as shown by the histogram above, although it appears as though there is an outlier. It is likely that the data would be more centrally distributed if this outlier was removed, but for the sake of exploring the dataset as it was provided to us, we opted to include the outlier.

```{r cases 2021 cor, include = FALSE}

cor1 <- cor(covid_healthcarecosts$cases_per_100k_2021, covid_healthcarecosts$avg_cost_2022)

cor1test <- cor.test(covid_healthcarecosts$cases_per_100k_2021, covid_healthcarecosts$avg_cost_2022)

```

```{r cases 2021 cor graph, echo = FALSE, results = TRUE}

ggplot(data = covid_healthcarecosts, aes(x=cases_per_100k_2021, y=avg_cost_2022)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red", size = 1.5) +  # Custom line of best fit
  annotate("text", x = 50, y = 1400, label = paste("Correlation:", round(cor1, 2), " (p-value:", round(cor1test$p.value, 3),")"),
           color = "black", size = 4, fontface = "bold") +
labs(title = "2021 COVID-19 Cases",
       x = "Average New COVID-19 Cases per Day per 100k",
       y = "Average Out of Pocket Healthcare Costs") + 
  theme_minimal() +
  theme(axis.text.x=element_text(face = "bold", size = 12), 
    axis.text.y = element_text(face = "bold", size = 12),  
    axis.title = element_text(face = "bold", size = 14),  
    panel.grid.major = element_line(color = "lightgray"),
    panel.background = element_rect(fill = "aliceblue"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

```
The correlation between average new COVID-19 cases per day per 100,000 people in 2021 and average annual out of pocket healthcare costs per person was `r round(cor1, 2)` which was significant at a p-value of 0.05 (p-value: `r round(cor1test$p.value, 3)`). The correlation coefficient reveals that there is a relatively weak negative association between these two variables. As you can see in the graph, there is one major outlier which has heavily impacted the direction of the relationship.

##### **Rank of Counties Based on 2021 COVID-19 Cases**

For the ranked variables, it is not necessary to view the distribution because only one county is assigned to each rank.

```{r cases 2021 ranked cor, include = FALSE}

cor3 <- cor(covid_healthcarecosts$rank2021case, covid_healthcarecosts$avg_cost_2022)
print(cor3)
cor3test <- cor.test(covid_healthcarecosts$rank2021case, covid_healthcarecosts$avg_cost_2022)
print(cor3test)

```

```{r cases 2021 ranked cor graph, echo = FALSE, results = TRUE}

ggplot(data = covid_healthcarecosts, aes(x=rank2021case, y=avg_cost_2022)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red", size = 1.5) +  # Custom line of best fit
  annotate("text", x = 50, y = 1400, label = paste("Correlation:", round(cor3, 2), " (p-value:", round(cor3test$p.value, 3),")"),
           color = "black", size = 4, fontface = "bold") +
labs(title = "2021 COVID-19 Cases",
       x = "County Rank",
       y = "Average Out of Pocket Healthcare Costs in US$") + 
  theme_minimal() +
 theme(axis.text.x = element_text(face = "bold", size = 12), 
    axis.text.y = element_text(face = "bold", size = 12),  
    axis.title = element_text(face = "bold", size = 14),  
    panel.grid.major = element_line(color = "lightgray"),
    panel.background = element_rect(fill = "aliceblue"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

```

The correlation between county ranks for the average new COVID-19 cases per day per 100,000 people in 2021 and average annual out of pocket healthcare costs per person was `r round(cor3, 2)` which was significant at a p-value of 0.05 (p-value: `r round(cor3test$p.value, 3)`). The correlation coefficient reveals that there is a relatively weak positive association between these two variables. As you can see in the graph, there is no longer a major outlier heavily influencing the direction of the relationship. A positive association makes more logical sense with the variables as it would be unlikely having more COVID-19 cases would decrease the out of pocket healthcare costs per person.

##### **New Daily COVID-19 Cases per 100K in 2022**

```{r cases 2022 summary, include = FALSE}

cases2022_sum <- summary(covid_healthcarecosts$cases_per_100k_2022)

```


```{r cases 2022 histogram, echo = FALSE, results = TRUE}

ggplot(data = covid_healthcarecosts, mapping = aes(x = cases_per_100k_2022)) +
  geom_histogram(fill = "white", color = "black", bins = 15) +
  labs(title = "Distribution of Average New COVID-19 Cases \nper Day per 100,000 People",
       subtitle = "Georgia, USA - 2022",
       x = "",
       y = "Number of Counties") +
  theme_minimal() +
  theme(axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold", size = 12),  # Larger y-axis labels
    axis.title = element_text(face = "bold", size = 14),  # Larger axis titles
    panel.grid.major = element_line(color = "lightgray"),  # Add light gray gridlines
    panel.background = element_rect(fill = "aliceblue"),  # Add light blue background
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

```

In 2022, across Georgia, the mean average new COVID-19 cases per day per 100,000 people for all of Georgia in 2022 was approximately `r round(cases2022_sum[4], 1)`, which is about 3 people less per 100,000 people than it was in 2021. There was a minimum of approximately `r round(cases2022_sum[1], 1)` new cases per day per 100,000 people, a maximum of approximately `r round(cases2022_sum[6], 1)` new cases per day per 100,000 people, and a median of approximately `r round(cases2022_sum[3], 1)`. The median is very slightly smaller than the mean, meaning there is a slight right-skew, although much like the 2021 data, there appears to be an outlier that is causing the skew. If the outlier were to be removed the data would be much more centrally distributed. Just as was the case for the 2022 case data, we opted to keep the data to have a better idea of the data's nature as it was given to us.

```{r cases 2022 cor, include = FALSE}

cor2 <- cor(covid_healthcarecosts$cases_per_100k_2022, covid_healthcarecosts$avg_cost_2022)
print(cor2)
cor2test <- cor.test(covid_healthcarecosts$cases_per_100k_2022, covid_healthcarecosts$avg_cost_2022)
print(cor2test)

```

```{r cases 2022 cor graph, echo = FALSE, results = TRUE}

ggplot(data = covid_healthcarecosts, aes(x=cases_per_100k_2022, y=avg_cost_2022)) +
  geom_point() +
 geom_smooth(method = "lm", se = FALSE, color = "red", size = 1.5) +  # Custom line of best fit
  annotate("text", x = 50, y = 1400, label = paste("Correlation:", round(cor2, 2), " (p-value:", round(cor2test$p.value, 3),")"),
           color = "black", size = 4, fontface = "bold") +
labs(title = "2022 COVID-19 cases",
       x = "Average New COVID-19 Cases per Day per 100k",
       y = "Average Out of Pocket Healthcare Costs") + 
  theme_minimal() +
  theme(axis.text.x = element_text(face = "bold", size = 12), 
    axis.text.y = element_text(face = "bold", size = 12),  
    axis.title = element_text(face = "bold", size = 14),  
    panel.grid.major = element_line(color = "lightgray"),
    panel.background = element_rect(fill = "aliceblue"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

```

The correlation between average new COVID-19 cases per day per 100,000 people in 2022 and average annual out of pocket healthcare costs per person was `r round(cor2, 2)` which was significant at a p-value of 0.05 (p-value: `r round(cor2test$p.value, 3)`). The correlation coefficient reveals that there is a relatively weak negative association between these two variables. Similarly to the 2021 cases, you can see in the graph that there is one major outlier which has heavily impacted the direction of the relationship.

##### **Rank of Counties Based on 2022 COVID-19 Cases**

```{r cases 2022 ranked cor, include = FALSE}

cor4 <- cor(covid_healthcarecosts$rank2022case, covid_healthcarecosts$avg_cost_2022)
print(cor4)
cor4test <- cor.test(covid_healthcarecosts$rank2022case, covid_healthcarecosts$avg_cost_2022)
print(cor4test)

```

```{r cases 2022 ranked cor graph, echo = FALSE, results = TRUE}

ggplot(data = covid_healthcarecosts, aes(x=rank2022case, y=avg_cost_2022)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red", size = 1.5) + 
  annotate("text", x = 50, y = 1400, label = paste("Correlation:", round(cor4, 2), " (p-value:", round(cor4test$p.value, 3),")"),
           color = "black", size = 4, fontface = "bold") +
labs(title = "2022 COVID-19 Cases",
       x = "County Rank",
       y = "Average Out of Pocket Healthcare Costs in US$") + 
  theme_minimal() +
   theme(axis.text.x = element_text(face = "bold", size = 12), 
    axis.text.y = element_text(face = "bold", size = 12),  
    axis.title = element_text(face = "bold", size = 14),  
    panel.grid.major = element_line(color = "lightgray"),
    panel.background = element_rect(fill = "aliceblue"),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5))

```

The correlation between county ranks for the average new COVID-19 cases per day per 100,000 people in 2022 and average annual out of pocket healthcare costs per person was `r round(cor4, 2)` which was significant at a p-value of 0.05 (p-value: `r round(cor4test$p.value, 3)`). The correlation coefficient reveals that there is a relatively weak positive association between these two variables. Similarly to the ranked cases in 2021, you can see in the graph that there is no longer a major outlier heavily influencing the direction of the relationship. Again, a positive association makes more logical sense with the variables as it would be unlikely having more COVID-19 cases would decrease the out of pocket healthcare costs per person.

---

### Mapping

Next, we wanted to see if there were geographic trends/centralized areas of high costs/cases in the state of Georgia. In order to do this we needed to merge our dataset with a shape file in order to visualize the data on a map. We utilized the [`urbnmapr`](https://github.com/UrbanInstitute/urbnmapr) package by the Urban Institute.

If there was a link between healthcare costs and COVID-19 cases per day, we would expect to see the same areas of high cases per day correspond with high or low annual healthcare costs, which should be apparent by similar color. We chose a yellow to red gradient to represent the increase in magnitude for either of our three variables, hoping that this gradient would elucidate if there existed a geographic trend wherein neighboring counties had similar magnitude case rates, healthcare costs, or both.

```{r map data}

counties1 <-counties %>% 
  mutate(fips=as.numeric(county_fips))
mapcovidcases <- left_join(covid_healthcarecosts, counties1, by=c("fips"))

```

```{r map total cases/costs, echo = FALSE, results = TRUE}

mapcovidcases %>%
  ggplot(aes(long, lat, group = group, fill = cases_per_100k_2021)) +
  geom_polygon() +
  scale_fill_gradient(low="yellow", high="red")+
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  labs(fill = "2021 Average cases per 100k per day")

mapcovidcases %>%
  ggplot(aes(long, lat, group = group, fill = cases_per_100k_2021)) +
  geom_polygon() +
  scale_fill_gradient(low="yellow", high="red")+
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  labs(fill = "2022 Average cases per 100k per day")

mapcovidcases %>%
  ggplot(aes(long, lat, group = group, fill = avg_cost_2022)) +
  geom_polygon(color=NA) +
  scale_fill_gradient(low="yellow", high="red")+
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  labs(fill = "2022 Average Annual out of pocket healthcare costs (USD$)")

```

However, it became evident that the high average daily case rate in Chattahoochee county was making this gradient incompatible with our design; since this county's case rate was much higher than every other county in Georgia, the gradient is unable to elucidate any geographic trends with any of our case data. Both 2021 and 2022 case variables have this problem, whereas the annual healthcare cost variable lacks an outlier to this extent, so its gradient reveals the possibility of a trend in higher healthcare costs in the Blue Ridge region of North Georgia (Northeast Corner) and possible trend of lower healthcare costs in Middle Georgia.
In order to evade the complication of Chattahoochee county's skewing observation without removing the data point from the analysis, we found it necessary to plot the ranks of each county in each variable. By doing this, we are able to see the same trend of whether a county has a higher magnitude observation for case rate or healthcare cost in comparison to neighboring counties. This makes it possible to see any geographic/variable trends between our variables, as this gradient would be able to reveal whether counties of higher ranks (closer to 1) are close to each other AND if high rank counties for daily COVID rates are also high rank counties for annual healthcare cost. This allows for us to keep Chattahoochee county's COVID daily rate data without jeopardizing our ability to view trends in other significantly lower magnitude daily case rate counties.

```{r map RANK, echo = FALSE, results = TRUE}

mapcovidcases %>%
  ggplot(aes(long, lat, group = group, fill = rank2021case)) +
  geom_polygon() +
  scale_fill_gradient(low="red", high="yellow")+
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  labs(fill = "2021 County rank of Average Cases per 100k per Day")

mapcovidcases %>%
  ggplot(aes(long, lat, group = group, fill = rank2022case)) +
  geom_polygon() +
  scale_fill_gradient(low="red", high="yellow")+
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  labs(fill = "2022 County rank of Average Cases per 100k per Day")

mapcovidcases %>%
  ggplot(aes(long, lat, group = group, fill = rank2022cost)) +
  geom_polygon(color=NA) +
  scale_fill_gradient(low="red", high="yellow")+
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  theme_void() +
  labs(fill = "2022 County rank of Annual Out of Pocket Healthcare Cost")
```

Now, we can see the gradient's intended effect of showing whether higher-ranking counties lie geographically close to one another and if there is a trend between high-ranking counties for average daily COVID-19 case rate and annual out-of-pocket healthcare costs. In 2022, we can see that the majority of high magnitude daily COVID 19 rate per 100k is centralized somewhat in Southwest GA, with the Blue Ridge region of North Georgia being mostly moderate to low magnitude daily COVID 19 rate per 100k in comparison to the rest of the state. In contrast, the high magnitude values for 2022 annual out-of-pocket healthcare costs seems to be centralized in the Blue Ridge region (in congruence with the previous map's analysis). 2021 seemed to have a centralization of high case rate in NW Georgia (somewhat present in 2021) and SE Georgia, which is distinct from the 2022 data. In general, there is not an positive geographic trend between out-of-pocket cost and COVID case rate elucidated by this analysis; in fact, there seems to be an inverse effect in the Blue Ridge region of North Georgia, though further statistical tests are required to see if this is indeed a statistical correlation.

By being able to see how each county compares with each other on a local and statewide scale, we have a better idea of the geographic trends present in Georgia counties for our three variables of interest. This preliminary visualization shows us how this data might be able to be used to reveal underlying policy/political reasons for any statistical correlation or lack thereof. 

---

### Correlation Analysis

In order to 

### Regression Analysis

We finally wanted to see if our COVID-19 cases variables are predictive of the average out of pocket healthcare costs per person. In order to do this we are utilizing multiple linear regression. Due to the outlier present in the non-ranked COVID-19 cases variables, the assumptions of multiple linear regression are violated. The outlier causes the residuals of the regression analysis to have non-constant variance as seen below.

```{r mlr, include=FALSE}

mlr <- lm(avg_cost_2022 ~ cases_per_100k_2021 * cases_per_100k_2022, data = covid_healthcarecosts)
summary(mlr)

residuals <- resid(mlr)
fitted_values <- fitted(mlr)

# Create the residual plot
ggplot(data = data.frame(fitted_values, residuals), aes(x = fitted_values, y = residuals)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Fitted values", y = "Residuals") +
  theme_minimal()

```

Since there is no evidence that the outliers in the data are due to human error, we cannot justify removing them to perform our analysis. Instead, we will utilize our ranked case variables and robust linear regression. Robust linear regression is less susceptible to issues regarding violations to the constant variance of residuals assumption, so it is more suitable for our data.

```{r rank regression set up, include=FALSE}

covid_healthcarecosts %>%
  mutate(rank2021case = factor(rank2021case),
         rank2022case = factor(rank2022case))

model_rank_robust <- lmrob(avg_cost_2022 ~ rank2021case + rank2022case, data = covid_healthcarecosts)
```


```{r rank regression plot, echo=FALSE, results=TRUE}

plot_summs(model_rank_robust)

```

Our regression analysis indicates that when rank of county increased by one in regard to the average COVID-19 cases in 2021, average annual out of pocket healthcare cost in 2022 increased by approximately 11 cents when keeping 2022 covid-19 case rank constant. The 2021 data was not statistically significant at a p-value of 0.05. On the other hand, when rank of county increased by one in regard to the average COVID-19 cases in 2022, average annual out of pocket healthcare cost in 2022 increased by approximately 48 cents when keeping 2021 covid-19 case rank constant. This was statistically significant at a p-value of 0.05, but not very relevant in actuality due to the increase in average annual out of pocket healthcare costs per person being by less than half a dollar for each rank increase. 

### References
| Barnett, M. L., Mehrotra, A., & Landon, B. E. (2020). Covid-19 and the upcoming financial crisis in health care. *NEJM Catalyst, 1*(2). 
|      [https://doi.org/10.1056/CAT.20.0153](https://doi.org/10.1056/CAT.20.0153).

| Cillizza, C. (2020, April 21). *Inside the Georgia governor's hugely high-stakes coronavirus gamble*. CNN.
|      [https://www.cnn.com/2020/04/21/politics/brian-kemp-georgia-reopen-coronavirus/index.html](https://www.cnn.com/2020/04/21/politics/brian-kemp-georgia-reopen-coronavirus/index.html).

| The Commonwealth Fund. (2023). *Georgia performance data*.
|      [https://www.commonwealthfund.org/datacenter/georgia#:~:text=%2344%20Georgia's%202022%20Rank,Georgia%20ranked%20%2344%20this%20year](https://www.commonwealthfund.org/datacenter/georgia#:~:text=%2344%20Georgia's%202022%20Rank,Georgia%20ranked%20%2344%20this%20year).

| Li, K., Al-Amin, M., & Rosko, M. D. (2023). Early financial impact of the COVID-19 pandemic on U.S. hospitals. *Journal of Healthcare Management *
|      */ American College of Healthcare Executives, 68*(4), 268-283. [https://doi.org/10.1097/JHM-D-22-00175](https://doi.org/10.1097/JHM-D-22-00175).

| Masterson, L. (2023, October 13). *The worst (and best) states for healthcare, ranked*. Forbes Advisor.
|      [https://www.forbes.com/advisor/health-insurance/best-worst-states-for-healthcare/](https://www.forbes.com/advisor/health-insurance/best-worst-states-for-healthcare/).

| McCain, J. L., Wang, X., Connell, K., & Morgan, J. (2022). Assessing the impact of insurance type on COVID-19 mortality in black and white
|      patients in the largest healthcare system in the state of Georgia. *Journal of the National Medical Association, 114*(2), 218-226.
|      [https://doi.org/10.1016/j.jnma.2022.01.002](https://doi.org/10.1016/j.jnma.2022.01.002).

| The New York Times. (2021). *Coronavirus (Covid-19) Data in the United States* [Data set]. The New York Times.
|      [https://github.com/nytimes/covid-19-data](https://github.com/nytimes/covid-19-data).

| Pike, J., Kompaniyets, L., Lindley, M. C., Saydah, S., & Miller, G. (2023). Direct medical costs associated with post-COVID-19 conditions among
|      privately insured children and adults. *Preventing Chronic Disease, 20*(220292). [http://dx.doi.org/10.5888/pcd20.220292](http://dx.doi.org/10.5888/pcd20.220292).

| PolicyMap. (2022). *Average annual dollars spent out of pocket per person on medical care in 2022* [Data set]. PolicyMap.
|      [https://www.policymap.com/](https://www.policymap.com/).

| Strochak, S., Williams, A. R., & Ueyama, K. (2019). *urbnmapr* [Data set]. Urban Institute. [https://github.com/UrbanInstitute/urbnmapr](https://github.com/UrbanInstitute/urbnmapr).

| United States Census Bureau. (2024). *Annual Resident Population Estimates, Estimated Components of Resident Population Change, and Rates*
|      *of the Components of Resident Population Change for States and Counties: April 1, 2020 to July 1, 2023* [Data set]. United States Census
|      Bureau. [https://www.census.gov/data/datasets/time-series/demo/popest/2020s-counties-total.html](https://www.census.gov/data/datasets/time-series/demo/popest/2020s-counties-total.html).

| Wanga, V., Chevinsky, J. R., Dimitrov, L. V., Gerdes, M. E., Whitfield, G. P., Bonacci, R. A., Nji, M. A. M., Hernandez-Romieu, A. C., Rogers-Brown,
|      J. S., McLeod, T., Rushmore, J., Lutfy, C., Bushman, D., Koumans, E., Saydah, S., Goodman, A. B., King, S. M. C., Jackson, B. R., & Cope, 
|      J. R. (2021). Long-term symptoms among adults tested for SARS-CoV-2-United States, January 2020-April 2021. *MMWR Morb Mortal Wkly,*
|      *70*(36), 1235-1241.
